---
title: "Assignment-2"
author: "Abhishek Kumar"
date: "22/10/2020"
output: html_document
---


#setting working Directory
```{r working directory}
setwd("/home/abhishek/Desktop/R assignment/Assignment 2")
```

## Task 1: Manipulation

# 1. Loading data from directory

```{r load data}

load("triathlon2019.Rdata")
dat = data.frame(triathlon) # Created dataframe data.
head(dat)
is.data.frame(dat)
dim(dat)
colnames(dat)

```

#  Explore data
```{r data exploration and cleaning}

# New column names
new_col_names = c("Race_number","Triathlete_name", "Country_code", "Gender", 
                  "Race_division", "Swim_time_InSecs", "TT_SwimToBike_InSecs"
                  ,"Bike_time_InSecs", "TT_BikeToRun_InSecs","Run_time_InSecs"
                  , "Total_time_InSecs" ) 

# Distance related to swim, bike and run

swim_distance_km = 3.862
bike_distance_km = 180.246
run_distance_km = 42.195
total_distance = swim_distance_km + bike_distance_km + run_distance_km

colnames(dat) = new_col_names # Updated column names to more understandable form.
head(dat)

# Removed rows with missing time.
# is.na(columns) returns true for the position which has NA value in Overall column 
dat = dat[is.na(triathlon$Overall) == FALSE, ] # Excluded position where value is False, which indicates NA
str(dat)
```


# 3. Appended columns swim_Speed, bike_speed and run_speed(in km/h)
```{r columnadded}
# Added columns for speed for swimming, biking and running by dividing the total
# distance traveled by total time taken in hr(converted using secs/3600).

dat$Swim_speed_kmhr = swim_distance_km/(dat$Swim_time_InSecs/3600)
dat$Bike_speed_kmhr = bike_distance_km/(dat$Bike_time_InSecs/3600)
dat$Run_speed_kmhr = run_distance_km/(dat$Run_time_InSecs/3600)

head(dat[,c("Swim_speed_kmhr", "Bike_speed_kmhr","Run_speed_kmhr")])
```

## Task 2: Analyzing/ Plotting

# 1,2 Used pair plot
```{r pairplot} 

colvec = c("red","blue")
names(colvec) = c("Male","Female")
pairs(dat[, c("Swim_speed_kmhr", "Swim_time_InSecs")], col = colvec[dat$Gender])
pairs(dat[,6:11], col = colvec[dat$Gender])

```

# 3 Overall time in transition
```{r}
#Total time calculated by adding the transition times for each athleted and then 
# Divided by overall time

total_transitiontime = apply(dat[,c("TT_SwimToBike_InSecs","TT_BikeToRun_InSecs")], 1, sum)
dat$total_transitiontime_percentage = (total_transitiontime/dat$Total_time_InSecs)*100
hist(dat$total_transitiontime_percentage, xlab = "Total transition time %", main = "TT time percentage for each athletes")

# From histogram we can see that most number of athletes had overall transition time in between 1% and 2%.

```

# 4. Histograms 

```{r histogram}

par(mfrow = c(1,3))
hist(dat$Swim_speed_kmhr, xlab = "Swim speed in km/hr", main = "Speed in km/hr")
hist(dat$Bike_speed_kmhr, xlab = "Bike speed in km/hr", main = "Speed in km/hr")
hist(dat$Run_speed_kmhr,  xlab = "Run speed in km/hr", main = "Speed in km/hr")
par(mfrow = c(1,1))

```

# 5. Athletes from each country

```{r tablecreation}
# From the plot we can observe that USA has maximum numbers of athletes.
table(dat$Country_code)
barplot(table(dat$Country_code), las = 2, xlab = "Countries", ylab = "Number of athletes", cex.names = 0.7)
```


# 6. Finishing Time for each division

```{r boxplot}
# Used split to group the data of column Total_time_InSecs for each athlete by the categorical values in Race_division, and used boxplot to plot finishing time for each categories.

par(mar = c(6,6,1,1), cex.axis = 0.7, mai=c(1,1,1,1))
boxplot(split(dat$Total_time_InSecs, dat$Race_division), xlab = "Race Divisions", ylab = "Finishing Time",main = "Boxplot",  las = 2)

```



# 7.  Arranged finishin time for each race division based on the lowest to highest median finishing time in each category.

```{r}

# Used sapply to get the median of finishing time in each category
# Applied order to get the list of indexes of ordered finishing time by median.
# Rearranged finishing_time_by_division by this ordered positions list and plotted the data.
par(mar = c(6,6,1,1), cex.axis = 0.7, mai=c(1,1,1,1))
finishing_time_by_division = split(dat$Total_time_InSecs/3600, dat$Race_division)
ord = order(sapply(finishing_time_by_division, median))
boxplot(finishing_time_by_division[ord], main ="Ordered finishing time by Race Division", xlab= "Race-divisions ", ylab = " Finishing Time", las=2)
```


# 8 Customized y axis with whole number labels in hr.

```{r}
# axis was used to give customized y-axis labels.
par(mar = c(6,6,1,1), cex.axis = 0.7, mai=c(1,1,1,1))
boxplot(finishing_time_by_division[ord], main ="Ordered finishing time by Race Division", xlab= "Race-divisions ", ylab = " Finishing Time", yaxt = "n", las = 2)
axis(2, at = seq(2,18, by = 1), labels = TRUE)

```


## Creativity

Plotted data of speeds and total time take by male and females respectively to complete the race.

```{r creativity}

# 609 females participated in the race in comparison to 1649 males.
# Used split to categorize finishing time for male and females.
# From the plots below we cans see that in all race types females were slower than the males
# overall females took longer time to complete the race.


colvec2 = c("Red", "Blue")

# Pie chart show distribution of males and females population.
pie(table(dat$Gender), col = colvec2)
legend("topleft", legend = c("Female","Male"), fill = colvec2, border=colvec2, cex =0.4)

par(mfrow = c(2,2))
boxplot(split(dat$Total_time_InSecs/3600, dat$Gender), col = colvec2, xlab = "Gender", ylab = "Finishing Time", main="Boxplot: Gender vs Finishing time", cex = 1)
legend("topleft", legend = c("Female","Male"), fill = colvec2, border=colvec2, cex =0.4)
boxplot(split(dat$Swim_speed_kmhr/3600, dat$Gender),col = colvec2, xlab = "Gender", ylab = "Swim speed", main="Boxplot: Gender vs Swim speed", cex = 1)
legend("topleft", legend = c("Female","Male"), fill = colvec2, border=colvec2, cex =0.4)
boxplot(split(dat$Bike_speed_kmhr/3600, dat$Gender),col = colvec2, xlab = "Gender", ylab = "Bike speed", main="Boxplot: Gender vs Bike speed", cex = 1)
legend("topleft", legend = c("Female","Male"), fill = colvec2, border=colvec2, cex =0.4)
boxplot(split(dat$Run_speed_kmhr/3600, dat$Gender), col = colvec2, xlab = "Gender", ylab = "Run speed", main="Boxplot: Gender vs Run speed", cex = 1)
legend("topleft", legend = c("Female","Male"), fill = colvec2, border=colvec2, cex =0.4)
par(mfrow = c(1,1))

```

